---
title: "The Economic and Public Health Impact of Severe Weather Events"
author: "Joe Pechacek"
date: "June 17, 2015"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Synopsis
The purpose of this analysis is to provide a high-level look at both the Economic 
consequences and Public Health impacts resulting from severe weather events.  The
data for this analysis comes from the U.S. National Oceanic and Atmospheric 
Administration (NOAA) storm database.  The Economic consequences for this analysis is 
measured in terms of Millions of Dollars of damage to both Property and Crops.
The impact on Public Health is a measure of Injuries and Fatalities as a result 
of severe weather events.  We will look at each impact group independently, though
there is an expectation that certain events will contribute to great loss in both 
Economic and Public Health measures such as Tornadoes or Flooding.

## Data Processing
The data for this analysis was provided through the Reproducible Research course 
website presented by Johns Hopkins University on Coursera.  The data can be 
downloaded at the following link:

 
[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)  


The raw data encompasses a range of observations from 1950 through November 2011 and 
represents a rather wide range of details over the 60 plus year period.  Prior to
January 1996 the data had no standardized format and event details appear to be 
at the discretion of the event report author.  Starting in 1996 there was an effort to 
improve the standardization of the data fields and variables as well as include a 
wider range of weather events.  Prior to 1996, the storm database consisted mostly 
of Tornadoes, Thunderstorms, Wind and Hail events.  Starting in 1996 the database 
recorded 48 standardized event types including events such as Hurricanes, Floods 
and Wildfires.  See the [NOAA Storm Events Database Details](http://www.ncdc.noaa.gov/stormevents/details.jsp)
for reference.

Given that 45 years of the database was limited to a small segment of severe weather
event types, it was deemed that a bias would exist for certain event types. For this
reason, the analysis focused on the more than 15 year segment since 1996 when a 
wider range of events were recorded. Using this subset of the database helps ensure
each event recorded since 1996 is represented in the database more equally and will
then reduce the impact of this bias.

### Loading and Preparation

Prior to conducting the analysis, the following steps were followed to load the
database into R and process the raw data to include the key data points for this 
analysis.  The raw data file for this analysis is compressed using BZ2 and should 
be saved to the current working directory of the R session.  To find the current 
working directory you can call the function `getwd()`.

Throughout the analysis there are a series of R packages used to make data analysis
run more smoothly and easily.  Prior to any data processing, the following packages 
should be loaded.

* knitr
* dplyr
* lubridate
* ggplot2
* tidyr

If you do not have these packages available on your local machine, they
should be installed first through [CRAN](cran.r-project.org) or a similar source.

```{r Load Packages}
require(knitr)
require(dplyr)
require(lubridate)
require(ggplot2)
require(tidyr)
```

With the required packages loaded, we first need to load the data using the `read.csv()` 
function.  It is fortunate that the `read.csv()` function already recognizes BZ2 
files, so the data can be read in directly.  This imported data will be stored to
a variable named "dataMaster" in order to keep an untouched version of the raw data
that can be called anytime for further analysis.

```{r Read Data}
dataMaster <- tbl_df(read.csv("repdata-data-StormData.csv.bz2", stringsAsFactors = FALSE))
```

Since the focus of this analysis is on severe weather events, a variable named "EVTYPE" 
is going to be of great interest.  There are 48 standard EVTYPEs defined by NOAA 
that should be used to ensure consistency.  To ensure we maintain this consistency
during this analysis, the EVTYPE variables will all be changed to Upper Case to 
reduce the potential of mis-matched case and a list of the NOAA standard event 
types is created for use later.

```{r EVTYPE}
dataMaster$EVTYPE <- toupper(dataMaster$EVTYPE)
StdEventNames <- toupper(c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood",
                           "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", 
                           "Drought", "Dust Devil", "Dust Storm", "Excessive Heat",
                           "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze",
                           "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", 
                           "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)",
                           "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning",
                           "Marine Hail", "Marine High Wind", "Marine Strong Wind", 
                           "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet",
                           "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado",
                           "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash",
                           "Waterspout", "Wildfire", "Winter Storm", "Winter Weather"))
```

The next step in the data preparation is to parse down the raw data to only those 
items that are of interest to our analysis. This will be placed into an object 
called "data" in order to preserve the "dataMaster" object created previously.  The 
following code segments will select those data points from 1996 to current; only 
those items that have an impact on Public Health (FATALITIES & INJURIES not equal 
to 0) and items that impact Economic consequences (PROPDMG & CROPDMG not equal to 0).
We should also change the PROPDMGEXP and CROPDMGEXP variables to class 'factor' 
as they represent a form of multiplier to be used for the values contained in the 
variables PROPDMG and CROPDMG.

```{r Parse Data}
data <- dataMaster %>%
     mutate(YEAR = year(mdy_hms(dataMaster$BGN_DATE))) %>%
     filter(YEAR > 1996)

data <- data[with(data, INJURIES != 0 | 
                          FATALITIES != 0 | 
                          PROPDMG != 0 | 
                          CROPDMG != 0), ]

data$PROPDMGEXP <- as.factor(data$PROPDMGEXP)
data$CROPDMGEXP <- as.factor(data$CROPDMGEXP)
```

This completes the loading and preparation of the data that can now be used for the
two areas of our analysis - impact on Public Health and impact on Economic consequences.
The final data frame object "data" will be the main data frame that will feed into
the follow up analysis.

#### Impact on Public Health

The first area of the analysis looked at is the impact on Public Health.  This 
impact can be assessed by using two variables, FATALITIES and INJURIES, and then
looking for patterns that will describe which severe weather events have the 
greatest impact.  This high-level look at which events have the greatest impact 
will simply determine the severe weather event type (EVTYPE) that has the greatest
number of FATALITIES and INJURIES combined.  The following code segments will 
step through the process followed for this analysis:

First we will create a subset of the data that only includes FATALITIES and 
INJURIES, create a new variable called FATAL_INJURED to show the combined total
of fatalities and injuries, then select only those variables that may be needed
for this analysis such as Year, State, County, Event Type along with the fatality
and injury variables.

```{r Health Data}
dataHealth <- data[with(data, INJURIES != 0 |
                             FATALITIES != 0), ]

dataHealth <- dataHealth %>%
     mutate(FATAL_INJURED = (FATALITIES + INJURIES)) %>%
     select(YEAR, STATE, COUNTYNAME, EVTYPE, FATALITIES, INJURIES, FATAL_INJURED)
```

Now that we have a subset just for the Public Health variables, we need to
standardize the event types (EVTYPE) to use the standard event types.  This will
be a two steps process as there are going to be certain events in the database
that will not follow the standard and must be corrected.  The first step is to 
identify all the EVTYPE variables that match one of the 48 standard types, and then a
strategy for assigning the remaining unmatched event types can be applied.  First
we need to match those that are already using the standard by identifying in a 
variable StdEVTYPE the standard event type that matches. Those that do not have 
a match will be set to NA.

```{r Match EVTYPE}
dataHealth <- dataHealth %>%
     mutate(StdEVTYPE = ifelse(EVTYPE %in% StdEventNames,
                               EVTYPE,
                               NA))
```

Upon inspection of the data during this analysis, it was found that there are
close to 80 event types that did not have a match to the 48 standard types.
Rather than correct all unmatched items, I looked at the impact these unmatched
items have on fatality and injury counts.  The strategy employed to correct these
unmatched events is that if a particular unmatched event had a small impact 
(for example, less than 10 fatalities or injuries), then it would not contribute
a great deal to the overall counts.  The following code was used to determine which
of the unmatched event types would have a large impact to this analysis by looking 
at a sample of 20 unmatched events that showed the largest count of fatalities 
and injuries.

```{r Unmatched Impact}
dataHealthTmp <- dataHealth %>%
     group_by(EVTYPE) %>%
     filter(is.na(StdEVTYPE)) %>%
     summarize(FATAL_INJURED = sum(FATAL_INJURED),
               FATALITIES = sum(FATALITIES),
               INJURIES = sum(INJURIES)) %>%
     arrange(desc(FATAL_INJURED))

head(as.data.frame(dataHealthTmp),20)

```

By inspection of this list, it can be seen that there are some unmatched events 
that had a large number of fatalities and injuries.  It can also be noted that 
this count drops quickly with item #20 "COLD"" having a count of only 27 
fatalities/injuries as compared to over 3,500 for item #1.  Based on this 
observation, I chose to only focus on matching those events that contribute more
than 50 total fatalities and injuries (items 1 - 16).  In order to match these
remaining event types, I first created a list of unmatched items and then used a 
series of `grep()` functions to assign the best match.  There will be additional
unmatched items that may match based on the `grep()` character string used, 
however these additional items are considered to be of minimal impact to the
scope based on the magnitude of the values.

```{r Match Unmatched}
EV_NAs <- data.frame(EVTYPE = unique(dataHealth$EVTYPE[is.na(dataHealth$StdEVTYPE)]),
                     stringsAsFactors = FALSE)
EV_NAs$StdEVTYPE <- NA

EV_NAs[grep("WIND", EV_NAs$EVTYPE), 2] <- "HIGH WIND"
EV_NAs[grep("FOG", EV_NAs$EVTYPE), 2] <- "DENSE FOG"
EV_NAs[grep("COLD", EV_NAs$EVTYPE), 2] <- "EXTREME COLD/WIND CHILL"
EV_NAs[grep("WINTER WEATHER", EV_NAs$EVTYPE), 2] <- "WINTER WEATHER"
EV_NAs[grep("WINTRY MIX", EV_NAs$EVTYPE), 2] <- "WINTER WEATHER"
EV_NAs[grep("TSTM", EV_NAs$EVTYPE), 2] <- "THUNDERSTORM WIND"
EV_NAs[grep("HURRICANE", EV_NAs$EVTYPE), 2] <- "HURRICANE (TYPHOON)"
EV_NAs[grep("FIRE", EV_NAs$EVTYPE), 2] <- "WILDFIRE"
EV_NAs[grep("GLAZE", EV_NAs$EVTYPE), 2] <- "WINTER WEATHER"
EV_NAs[grep("RIP", EV_NAs$EVTYPE), 2] <- "RIP CURRENT"
EV_NAs[grep("SURF", EV_NAs$EVTYPE), 2] <- "HIGH SURF"
EV_NAs[grep("LANDSLIDE", EV_NAs$EVTYPE), 2] <- "DEBRIS FLOW"
EV_NAs[grep("STREAM FLD", EV_NAs$EVTYPE), 2] <- "FLOOD"
```

With the list of updated EVTYPEs assigned to one of the 48 standard types, we can
now merge this list with our current data object.  The `left_join()` process will 
combine the data and create two new variables that will need to be combined into
a single variable StdEVTYPE.
```{r Join Data}
dataHealth <- left_join(dataHealth, EV_NAs, by = "EVTYPE")

dataHealth <- dataHealth %>%
     mutate(StdEVTYPE = ifelse(is.na(dataHealth$StdEVTYPE.x),
                                  dataHealth$StdEVTYPE.y,
                                  dataHealth$StdEVTYPE.x)) %>%
     select(-StdEVTYPE.x, -StdEVTYPE.y)
```

With this updated data object that now includes updates to use the 48 standard 
event types for events with a greater impact, we can focus on performing the 
analysis to answer the question
**"Across the United States, which types of events are most harmful with respect to population health?"** 
The first step I took was to exclude the remaining event types that were not 
matched earlier.  With this resulting data set, I sorted the data based on the
greatest number of Fatalities and Injuries combined. This is then used to create
a plotting data set that only includes the 10 events with the greatest impact to 
Public Health.  This plotting data set is converted to a long format using 
`gather()` to aid in the plotting function that will create separate colors for
Fatalities and Injuries.

```{r Plot Data 1}
DH_Analysis <- dataHealth %>%
     group_by(StdEVTYPE) %>%
     filter(!is.na(StdEVTYPE)) %>%
     summarize(FATAL_INJURED = sum(FATAL_INJURED),
               FATALITIES = sum(FATALITIES),
               INJURIES = sum(INJURIES)) %>%
     arrange(desc(FATAL_INJURED))

DH_plot_1 <- DH_Analysis[1:10,]
DH_plot_1 <- DH_plot_1 %>%
     select(StdEVTYPE, FATALITIES, INJURIES) %>%
     gather(Measurement, Value, -StdEVTYPE) %>%
     group_by(Measurement)
```

Using the resulting data, we can generate a plot using the `ggplot2` package.
The results of this plot will be discussed further in the **Results** section 
later in this report.

```{r Plot 1}
p1 <- ggplot(DH_plot_1, aes(x = reorder(StdEVTYPE, Value, function(x)max(x)),
                            y = Value, fill = Measurement)) +
     geom_bar(stat = "identity") +
     coord_flip() +
     labs(title = "Event Types with Greatest Overall Impact on Public Health since 1996",
          x = "Event Type",
          y = "Number of Fatalities and Injuries")
```


#### Impact on Economic Factors

This area of the analysis focuses on the Economic consequences of severe weather.
This impact will be focused on two key variables CROPDMG and PROPDMG along
with their requisite scalars CROPDMGEXP and PROPDMGEXP.  

Most of the processing flows used here are similar to those previously used in 
the Public Health analysis, so descriptions of each step will be less detailed, 
however key variables and decisions made will be of equal detail as used in the 
Public Health analysis.

As we did in the Public Health analysis, we will create a new data object that is
now going to be focused on the CROPDMG and PROPDMG variables, then the process to 
match the EVTYPE to one of the 48 standard event types is completed.

```{r Econ Data}
dataEcon <- data[with(data, PROPDMG != 0 |
                             CROPDMG != 0), ]

dataEcon <- dataEcon %>%
     mutate(StdEVTYPE = ifelse(EVTYPE %in% StdEventNames,
                               EVTYPE,
                               NA))

```

The next step in the analysis is to determine which of the unmatched events
have a significant impact on the overall analysis.  This will follow a similar
process as used previously with Public Health, however the variables in this 
case are in terms of damage costs in Dollars.  Before we can run any comparisons
of damage costs, we must first put all damage costs into the same unit of measure.
The database makes use of the scalar variables CROPDMGEXP and PROPDMGEXP for the 
cost of damages.  This scalar is signified by "K" = "Thousand", "M" = "Million" 
or "B" = "Billion".  In order to compare the impact of each event, the total 
dollar amount for the CROPDMG and PROPDMG variables needs to be calculated.  

The first step in this process is to update the CROPDMGEXP and PRODMGEXP variables
to be numeric representations of the scale factors.  Given the magnitude of the 
values, I have adjusted the multiplier to result in an outcome of Millions 
(i.e. Billion = 1000, Thousand = 0.001).

With these scale factors in a numeric format, a new variable can be created that 
shows the total impact in Millions of Dollars.  After this conversion of total
Dollar impact is completed, the scale factor variables are not needed and removed
from the data set.
```{r Dollar Scale}
dataEcon$PROPDMGEXP <- sub("K", 0.001, dataEcon$PROPDMGEXP)
dataEcon$PROPDMGEXP <- sub("M", 1, dataEcon$PROPDMGEXP)
dataEcon$PROPDMGEXP <- sub("B", 1000, dataEcon$PROPDMGEXP)
dataEcon$PROPDMGEXP <- sub("", 0, dataEcon$PROPDMGEXP)
dataEcon$CROPDMGEXP <- sub("K", .001, dataEcon$CROPDMGEXP)
dataEcon$CROPDMGEXP <- sub("M", 1, dataEcon$CROPDMGEXP)
dataEcon$CROPDMGEXP <- sub("B", 1000, dataEcon$CROPDMGEXP)
dataEcon$CROPDMGEXP <- sub("", 0, dataEcon$CROPDMGEXP)

class(dataEcon$PROPDMGEXP) <- "numeric"
class(dataEcon$CROPDMGEXP) <- "numeric"


dataEcon <- dataEcon %>%
     mutate(PROPDMG_M = PROPDMG * PROPDMGEXP, 
            CROPDMG_M = CROPDMG * CROPDMGEXP,
            TOTALDMG_M = CROPDMG_M + PROPDMG_M) %>%
     select(-CROPDMGEXP, -PROPDMGEXP, -PROPDMG, -CROPDMG)
```

With a total damage amount for each event in a common unit of measure, the next
step is to go through the matching process.  We will use the same method of 
identifying the unmatched event types with the greatest impact and then update a
list of unmatched events to be merged later. From the resulting list below, a 
determination can be made on which unmatched event types should be considered as 
significant.  In this case, we can once again see that there is a quick decline 
in the values associated with unmatched events and I chose events with over $50 
Million in damage as being significant for this analysis.  Based on this threshold, 
I used `grep()` to match select event types to one of the 48 standard types. As before
it is expected that some lesser events may also be updated based on the `grep()` 
character string used and this was considered acceptable.

``` {r Unmatched Events}
dataEconTmp <- dataEcon %>%
     group_by(EVTYPE) %>%
     filter(is.na(StdEVTYPE)) %>%
     summarize(PROPDMG_M = sum(PROPDMG_M),
               CROPDMG_M = sum(CROPDMG_M),
               TOTALDMG_M = sum(TOTALDMG_M)) %>%
     arrange(desc(TOTALDMG_M))

head(as.data.frame(dataEconTmp),20)


EV_NAs <- data.frame(EVTYPE = unique(dataEcon$EVTYPE[is.na(dataEcon$StdEVTYPE)]),
                     stringsAsFactors = FALSE)
EV_NAs$StdEVTYPE <- NA


EV_NAs[grep("HURRICANE", EV_NAs$EVTYPE), 2] <- "HURRICANE (TYPHOON)"
EV_NAs[grep("TSTM", EV_NAs$EVTYPE), 2] <-  "THUNDERSTORM WIND"
EV_NAs[grep("SURGE", EV_NAs$EVTYPE), 2] <-  "STORM SURGE/TIDE"
EV_NAs[grep("FIRE", EV_NAs$EVTYPE), 2] <-  "WILDFIRE"
EV_NAs[grep("COLD", EV_NAs$EVTYPE), 2] <-  "EXTREME COLD/WIND CHILL"
EV_NAs[grep("TYPHOON", EV_NAs$EVTYPE), 2] <- "HURRICANE (TYPHOON)"
EV_NAs[grep("LANDSLIDE", EV_NAs$EVTYPE), 2] <-  "DEBRIS FLOW"
EV_NAs[grep("FREEZE", EV_NAs$EVTYPE), 2] <-  "FROST/FREEZE"
EV_NAs[grep("COASTAL FLOODING", EV_NAs$EVTYPE), 2] <-  "COASTAL FLOOD"
EV_NAs[grep("STREAM FLD", EV_NAs$EVTYPE), 2] <-  "FLOOD"
EV_NAs[grep("FLOOD", EV_NAs$EVTYPE), 2] <- "FLOOD"
EV_NAs[grep("HAIL", EV_NAs$EVTYPE), 2] <-  "HAIL"
```

Now that we have updated event types, the final steps in this analysis are to
merge the data to include these updates; create an analysis data set sorted on
events with the highest damage cost; then prepare a plotting data set in a
long format using `gather()` to include the 10 events with the highest total 
damage costs.

```{r Join Plot Data 2}
dataEcon <- left_join(dataEcon, EV_NAs, by = "EVTYPE")

dataEcon <- dataEcon %>%
     mutate(StdEVTYPE = ifelse(is.na(dataEcon$StdEVTYPE.x),
                                  dataEcon$StdEVTYPE.y,
                                  dataEcon$StdEVTYPE.x)) %>%
     select(-StdEVTYPE.x, -StdEVTYPE.y)


DE_Analysis <- dataEcon %>%
     group_by(StdEVTYPE) %>%
     filter(!is.na(StdEVTYPE)) %>%
     summarize(TOTALDMG_M = sum(TOTALDMG_M),
               PROPDMG_M = sum(PROPDMG_M),
               CROPDMG_M = sum(CROPDMG_M)) %>%
     arrange(desc(TOTALDMG_M))

DE_plot_1 <- DE_Analysis[1:10,]
DE_plot_1 <- DE_plot_1 %>%
     select(StdEVTYPE, CROPDMG_M, PROPDMG_M) %>%
     gather(Measurement, Value, -StdEVTYPE) %>%
     group_by(Measurement)
```

With this plotting data set completed, we can generate a plot that helps to answer the
question: 
**"Across the United States, which types of events have the greatest economic consequences?"**  Again the `ggplot2` package is used for this.

```{r Plot 2}
p2 <- ggplot(DE_plot_1, aes(x = reorder(StdEVTYPE, Value, function(x)max(x)),
                            y = Value, fill = Measurement)) +
     geom_bar(stat = "identity") +
     coord_flip() +
     labs(title = "Event Types with Greatest Overall Economic Consequences since 1996",
          x = "Event Type",
          y = "Millions of Dollars in Damage")
```


## Results

These two analyzes addressed two rather high-level questions about total impact
across the United States for either Public Health or Economic consequences.  Based
on this high-level analysis, further analysis in specific areas may be considered
to address more specific questions or to focus on specific relationships.  For 
example, the impact of Tornadoes may be seen in a variety of states, but there are
certain states in the Midwest that are most likely to have Tornadoes.  Additionally,
Hurricanes are going to be focused on coastal areas, so a more detailed analysis 
of areas most impacted could be another consideration.  This analysis made no 
attempt to go beyond the initial scope of identifying events with the greatest 
impact across the United States, hence the following results are based on this 
high-level view.  

#### Public Health impact

With respect to Public Health, the following chart shows that Tornadoes are by far
the greatest contributor to a combined count of injuries and fatalities. The events
with the next highest impact are less than half that of Tornadoes.

```{r Print Plot 1}
print(p1)
```

The relationship between fatalities and injuries should also be considered as
it is clear that injuries are of the greatest impact.  Also note that the total 
impact of both fatalities and injuries for the 10th highest event type (High Wind)
is nearly the same or less than fatalities for Tornadoes and Excessive Heat events.
This relationship provides validation that the results identified those events with 
the highest counts in fatalities and injuries as there are events with higher counts in 
one aspect (fatalities) than the total count of fatalities and injuries seen in other
lower ranked events. 

#### Economic Consequences

With respect to Economic consequences, the following chart shows that Flood 
is by far the greatest impact to combined damage costs.  Again we can see that
the #2 ranked event Hurricane (Typhoon) is almost half the impact as #1 Flood.

```{r Print Plot 2}
print(p2)
```

Also note that the Crop damage is low for all types except Drought which may be
expected.  The fact that Drought is one of the events in this list provides 
further validation of the results as it provides an event with high damage costs
in one aspect (Crops) that is greater than the total damage costs for some lower 
ranked events.

### Summary

By comparison of the two charts presented above, it can be seen that Flood 
presents a great impact on both Public Health and Economic consequences as it
ranks #1 or #2 in each analysis.  It is also interesting to note that the following
`r length(intersect(DH_plot_1$StdEVTYPE, DE_plot_1$StdEVTYPE))` 
event types are included in the Top 10 for both lists.

```{r Intersect EVTYPE}
intersect(DH_plot_1$StdEVTYPE, DE_plot_1$StdEVTYPE)
```

Further analysis may be warranted for the event types identified during this 
analysis, and in particular the 
`r length(intersect(DH_plot_1$StdEVTYPE, DE_plot_1$StdEVTYPE))`
events that fall into the Top 10 of both analyses. Conducting further analysis may 
provide more details on any trends and/or impacts based on other factors such 
as geographic location (Coastal versus Inland) or possibly if there are connections
to population density (Metropolitan versus Rural).  This analysis was not designed 
nor intended to provide this level of detail.
